Calendar-based graphics for visualizing people's daily schedules

### 2017-07-29 main.Rmd
small changes to intro

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly interval by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

We develop a method for organizing and visualizing temporal data, collected on sub-daily interval, into a calendar layout. We hope this calendar-based visualization to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily rhythms, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides daily and sub-daily patterns by faceting on day of the week. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

@@

@@

A collection of data plots, organized in a familiar format, offer an intuitive and contextual way to understand complex data. @Wickham2012glyph embedded a sequence of daily temperature line charts into a glyph map according to the spatial locations; and @R-geofacet provided methods in the \pkg{geofacet} package to arrange data plots into a grid, while preserving the geographical layout. They both attempt to enable one to see the spatial context of each self-contained data plot.

Alternatively, calendar-based graphics turn out to be a useful tool in unfolding human-related activities over time. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasons like school vacations; all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than day. Time of day, which serves as one of the most important aspects in explaining substantial variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using the curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout, (3) to enable an overlaying plot consisting of multiple time series. The proposed algorithm has been implemented in the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

We develop a method for organizing and visualizing temporal data, collected on sub-daily interval, into a calendar layout. The purpose of the calendar-based visualization is to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area (Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily rhythms, to assist the city administration and local businesses with event planning and operational management. A routine examination of the data would be constructing a conventional time series plot to catch a glimpse of temporal patterns. The facetted plots in Figure \ref{fig:time-series-plot}, give an overall picture of the foot traffic at 3 different sensors over 2016. Further facetting by day of the week (Figure \ref{fig:facet-time}) provides a better glimpse at the daily and sub-daily, like commuter, patterns. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

@@

and the conventional ways to display time series and constructing facetted displays don't completely address these challenges.

@@

The work is inspired by @Wickham2012glyph, which uses linear algebra to display spatio-temporal data as glyphs on maps.  It is also related to recent work by
@R-geofacet which provides methods in the \pkg{geofacet} package to arrange data plots into a grid, while preserving the geographical position. Both of these show data in a spatial context.

In constrast, calendar-based graphics unpacks the temporal variable, at different resolutions, to digest multiple seasonalities, and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasons like school vacations; all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout, (3) to enable an overlaying plots consisting of multiple time series. The proposed algorithm has been implemented in the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

### 2017-07-29 main.tex
small changes to intro

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly interval by sensors scattered around the city.
Faceting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format options and variations. It is implemented in the R
package \pkg{sugrrants}.
}

@@

We develop a method for organizing and visualizing temporal data,
collected on sub-daily interval, into a calendar layout. We hope this
calendar-based visualization to provide insights into people's daily
schedules relative to events such as work days, weekends, holidays, and
special events. This work was originally motivated by studying foot
traffic in the city of Melbourne \citep{ped}. There have been 43 sensors
installed that count pedestrians every hour across the downtown area up
until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can
shed light into understanding people's daily rhythms, or assisting
administration and business planning. We start off with the conventional
time series plot to catch a glimpse of such data. A small multiples,
shown in Figure \ref{fig:time-series-plot}, gives an overall picture of
the foot traffic at 3 different sensors. Figure \ref{fig:facet-time}
provides daily and sub-daily patterns by faceting on day of the week.
The sensor data, like many temporal datasets on human behaviors, lends
itself to a number of exploratory data visualization challenges:

@@

A collection of data plots, organized in a familiar format, offer an
intuitive and contextual way to understand complex data.
\citet{Wickham2012glyph} embedded a sequence of daily temperature line
charts into a glyph map according to the spatial locations; and
\citet{R-geofacet} provided methods in the \pkg{geofacet} package to
arrange data plots into a grid, while preserving the geographical
layout. They both attempt to enable one to see the spatial context of
each self-contained data plot.

Alternatively, calendar-based graphics turn out to be a useful tool in
unfolding human-related activities over time. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasons like school vacations; all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in two R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are limited to
colour-encoding graphics and are unable to use time scales smaller than
day. Time of day, which serves as one of the most important aspects in
explaining substantial variations arisen from pedestrian sensor data,
will be neglected through daily aggregation. Additionally if simply
using colored blocks rather than curves, it may become perceptually
difficult to estimate the shape positions and changes, although using
the curves comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to
incorporate line graphs and other types of glyphs into the graphical
toolkit for the calendar layout, (3) to enable an overlaying plot
consisting of multiple time series. The proposed algorithm has been
implemented in the \code{frame_calendar} function in the \pkg{sugrrants}
package \citep{R-sugrrants} using \proglang{R} \citep{R-base}.

>>>>

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly intervals by sensors scattered around the city.
Faceting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format options and variations. It is implemented in the R
package \pkg{sugrrants}.
}

@@

We develop a method for organizing and visualizing temporal data,
collected on sub-daily interval, into a calendar layout. The purpose of
the calendar-based visualization is to provide insights into people's
daily schedules relative to events such as work days, weekends,
holidays, and special events. This work was originally motivated by
studying foot traffic in the city of Melbourne \citep{ped}. There have
been 43 sensors installed that count pedestrians every hour across the
downtown area (Figure \ref{fig:ped-map}). The dataset can shed light
into understanding people's daily rhythms, to assist the city
administration and local businesses with event planning and operational
management. A routine examination of the data would be constructing a
conventional time series plot to catch a glimpse of temporal patterns.
The facetted plots in Figure \ref{fig:time-series-plot}, give an overall
picture of the foot traffic at 3 different sensors over 2016. Further
facetting by day of the week (Figure \ref{fig:facet-time}) provides a
better glimpse at the daily and sub-daily, like commuter, patterns. The
sensor data, like many temporal datasets on human behaviors, lends
itself to a number of exploratory data visualization challenges:

@@

The work is inspired by \citet{Wickham2012glyph}, which uses linear
algebra to display spatio-temporal data as glyphs on maps. It is also
related to recent work by \citet{R-geofacet} which provides methods in
the \pkg{geofacet} package to arrange data plots into a grid, while
preserving the geographical position. Both of these show data in a
spatial context.

In constrast, calendar-based graphics unpacks the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasons like school vacations; all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in two R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are limited to
colour-encoding graphics and are unable to use time scales smaller than
day. Time of day, which serves as one of the most important aspects in
explaining substantial variations arising from pedestrian sensor data,
will be neglected through daily aggregation. Additionally, if simply
using colored blocks rather than curves, it may become perceptually
difficult to estimate the shape positions and changes, although using
curves comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to
incorporate line graphs and other types of glyphs into the graphical
toolkit for the calendar layout, (3) to enable an overlaying plots
consisting of multiple time series. The proposed algorithm has been
implemented in the \code{frame_calendar} function in the \pkg{sugrrants}
package \citep{R-sugrrants} using \proglang{R} \citep{R-base}.
